Cover Your Basis: Comprehensive Data-Driven Characterization of the Binary Black Hole Population

### 2022-10-17 src/tex/methods.tex
remove text stating dif min bH masses -- all min mass =5

We parameterize each binaries' masses with the primary (more massive component) mass ($m_1$) 
and the mass ratio ($q=m_2/m_1$) with support from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger: component spin magnitudes $a_1$ and $a_2$, and (cosine of) the tilt angles of each component, $\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binary's orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. For the primary mass distribution, we model the log probability with a B-Spline interpolated over 
knots linearly spaced in $\log(m_1)$ in the range of support given our observations ($[6.5\msun-100\msun]$), such that $\log(p(m_1 | \bm{c})) \propto B_{k}(\log(m_1) | \bm{c})$, where $B_{k}$ is the 
$k$-th order B-Spline function with basis coefficients $\bm{c}$. We follow the same procedure for the models in mass ratio and spin distributions with knots spaced linearly across each domain 
so that we have $\log(p(\theta | \bm{c}_\theta)) \propto B_{k}(\theta | \bm{c}_\theta)$, where $\theta$ can be $q$, $a_1$, $a_2$, $\cos{\theta_1}$ or $\cos{\theta_2}$. 
For the spin magnitude and tilt distributions we construct two versions of the model: first, we model  
each component's distribution as independently and identically distribution (IID), where we have a single B-Spline model and parameters (coefficients) for each binary spin. 
Secondly, we model each component's distribution to be unique, fitting separate sets of coefficients for the B-Spline models of the primary and secondary spin distributions. 
Lastly, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and its redshift, which we now use interchangeably. 
We take a semi-parametric approach to model the evolution of the merger rate with redshift, following \citet{Edelman_2022ApJ}, that parameterizes modulations to an underlying model 
with splines (in our case basis splines). We use the \textsc{PowerlawRedshift} model as the underlying distribution to modulate, which has a single hyperparameter, $\lambda_z$, and 
probability density defined as: $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. For more detailed descriptions of each model and 
specific prior choices used for the hyperparameters see Appendix \ref{sec:modelpriors}. Now that we have our comprehensive data-driven population model over 
each of the 7 modeled parameters built, we fit our model using the usual hierarchical Bayesian inference framework (see appendix \ref{sec:hierarchical_inference} for a review) in order to characterize 
the population of BBHs in the most recent catalog of GW observations, GWTC-3 \citep{GWTC3,o3b_astro_dist}.

>>>>

We parameterize each binaries' masses with the primary (more massive component) mass ($m_1$) 
and the mass ratio ($q=m_2/m_1$) with support from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger: component spin magnitudes $a_1$ and $a_2$, and (cosine of) the tilt angles of each component, $\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binary's orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. For the primary mass distribution, we model the log probability with a B-Spline interpolated over 
knots linearly spaced in $\log(m_1)$ from a minimum black hole mass, which we fix to $5\msun$, and a maximum mass 
that we set to $100\msun$. We then have the hyper-prior on primary mass with log probability density 
$\log(p(m_1 | \bm{c})) \propto B_{k}(\log(m_1) | \bm{c})$, where $B_{k}$ is the 
$k$-th order B-Spline function with basis coefficients $\bm{c}$. We follow the same procedure 
for the models in mass ratio and spin distributions with knots spaced linearly across each domain 
so that we have $\log(p(\theta | \bm{c}_\theta)) \propto B_{k}(\theta | \bm{c}_\theta)$, 
where $\theta$ can be $q$, $a_1$, $a_2$, $\cos{\theta_1}$ or $\cos{\theta_2}$. 
For the spin magnitude and tilt distributions we construct two versions of the model: first, we model  
each component's distribution as independently and identically distribution (IID), 
where we have a single B-Spline model and parameters (coefficients) for each binary spin. 
Secondly, we model each component's distribution to be unique, fitting separate sets of coefficients for the B-Spline models of the primary and secondary spin distributions. 
Lastly, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and its redshift, which we now use interchangeably. 
We take a semi-parametric approach to model the evolution of the merger rate with redshift, following \citet{Edelman_2022ApJ}, that parameterizes modulations to an underlying model 
with splines (in our case basis splines). We use the \textsc{PowerlawRedshift} model as the underlying distribution to modulate, which has a single hyperparameter, $\lambda_z$, and 
probability density defined as: $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. For more detailed descriptions of each model and 
specific prior choices used for the hyperparameters see Appendix \ref{sec:modelpriors}. Now that we have our comprehensive data-driven population model over 
each of the 7 modeled parameters built, we fit our model using the usual hierarchical Bayesian inference framework (see appendix \ref{sec:hierarchical_inference} for a review) in order to characterize 
the population of BBHs in the most recent catalog of GW observations, GWTC-3 \citep{GWTC3,o3b_astro_dist}.

### 2022-10-17 src/tex/results.tex
remove text stating dif min bH masses -- all min mass =5

\begin{figure*}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/mass_distribution_plot.pdf}
        \caption{The marginal primary mass distribution inferred with the B-Spline model (red), with 64 knots spaced linearly in $\log m_1$, from 6.5\msun 
        to 100\msun. The solid line shows the population predictive distribution (PPD), and the shaded region 
        the 90\% credible interval. We show the inferred PPD from the \textsc{PowerlawPeak} (blue) and \textsc{PowerlawSpline} (green) models 
        from the LVK's GWTC-3 population analyses \citep{o3b_astro_dist}.}
        \label{fig:mass_distribution}
    \end{centering}
    \script{mass_distribution_plot.py}
\end{figure*}

@@

Figure \ref{fig:mass_distribution} shows the primary mass distribution inferred with our B-Spline model (red), where we see features consistent with those inferred by the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \citep{Talbot_2018,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our B-Spline model finds peaks in merger rate density as a function of primary mass at both $\sim10\msun$ and $\sim35\msun$, agreeing with those 
reported using the same dataset in \citet{o3b_astro_dist}. The B-Spline model finds the same feature at $\sim18\msun$ as the \textsc{PowerlawSpline} model, but remains consistent with the \textsc{PowerlawSpline} model; the mass distribution is more uncertain in this region.
We find the largest disagreement at low masses, where the power-law-based models show a higher rate below ${\sim}8-9\msun$. This is partly due to the minimum mass hyperparameter (where the power law ``begins'') serving as the minimum allowable primary and secondary masses of the catalog. This leads to inferences of $m_\mathrm{min}$ below the minimum observed primary mass in the catalog, $\sim 6.4\msun$, to account for secondary BBH masses 
that can possibly get as low as $\sim 3\msun$. We choose a minimum primary mass for 
the B-Spline model to be $6.5\msun$, while secondary masses are enforced to be greater than $3\msun$. The lack of observations of binaries with low primary mass make rate estimates in this region strongly model and prior dependent.  We could be seeing signs of a decrease in merger rate from a ``lower mass gap'' between neutron star and BH masses, or we could be seeing fluctuations due to low-number statistics.  Either way we expect this to be resolved with future catalog updates. We also find no evidence for a sharp fall off in merger rate either following 
the pileup at ${\sim}35\,\msun$ -- expected if such a pileup was due to pulsational pair instability supernovae (PPISNe) -- or where the maximum mass truncation of the power law models 
are inferred. The lack of any high mass truncation, along with the peak at $\sim35\msun$ (significantly lower than expected from PPISNe) may pose challenges for conventional stellar 
evolution theory. This could be hinting at the presence of subpopulations that avoid pair instability supernovae during binary formation, but the confirmation of the existence of such subpopulations 
cannot be determined with the current catalog.

\begin{figure}[h]
    \includegraphics[width=\linewidth]{figures/q_distribution_plot.pdf}
    \caption{The marginal mass ratio distribution inferred with the B-Spline model (red), with 18 knots spaced linearly in $q$, from 0.03 to 1. The solid line shows the population predictive distribution (PPD), and the shaded region the 90\% credible interval. 
    We show the inferred PPD from the \textsc{PowerlawPeak} (blue) and \textsc{PowerlawSpline} (green) models from the LVK's GWTC-3 population analyses \citep{o3b_astro_dist}.}
    \label{fig:q_distribution}
    \script{q_distribution_plot.py}
\end{figure}

The marginal mass ratio distribution inferred by the B-spline model is shown in figure \ref{fig:q_distribution}.  These results suggest we may be seeing the first signs of departure from a simple power law behavior.  We find a potential plateau or decrease in the merger rate near equal mass ratios, as well as a broader tail towards unequal mass ratios 
than the power law based models find. Our results also suggest a shallower slope from $q\sim0.3$ to $q\sim0.7$, though uncertainty is larger in this region. The sharp decrease in rate just below 
$q\sim0.3$ is due to the minimum mass ratio truncation defined by $q_\mathrm{min}=\frac{m_\mathrm{min}}{m_1}$. When marginalizing over the primary mass 
distribution with a strong peak at $10\msun$, the mass ratio distribution truncates at $q\sim0.3$: the minimum secondary mass, $3\msun$, 
divided by the most common primary mass, ${\sim}10\msun$.

>>>>

\begin{figure*}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/mass_distribution_plot.pdf}
        \caption{The marginal primary mass distribution inferred with the B-Spline model (red), with 64 knots spaced linearly in $\log m_1$, from 5\msun 
        to 100\msun. The solid line shows the population predictive distribution (PPD), and the shaded region 
        the 90\% credible interval. We show the inferred PPD from the \textsc{PowerlawPeak} (blue) and \textsc{PowerlawSpline} (green) models 
        from the LVK's GWTC-3 population analyses \citep{o3b_astro_dist}.}
        \label{fig:mass_distribution}
    \end{centering}
    \script{mass_distribution_plot.py}
\end{figure*}

@@

Figure \ref{fig:mass_distribution} shows the primary mass distribution inferred with our B-Spline model (red), where we see features consistent with those inferred by the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \citep{Talbot_2018,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our B-Spline model finds peaks in merger rate density as a function of primary mass at both $\sim10\msun$ and $\sim35\msun$, agreeing with those 
reported using the same dataset in \citet{o3b_astro_dist}. The B-Spline model finds the same feature at $\sim18\msun$ as the \textsc{PowerlawSpline} model, but remains consistent with the \textsc{PowerlawSpline} model; the mass distribution is more uncertain in this region.
We find the largest disagreement at low masses, where the power-law-based models show a higher rate below ${\sim}8-9\msun$. This is partly due to the minimum mass hyperparameter (where the power law ``begins'') serving as the minimum allowable primary and secondary masses of the catalog. This leads to inferences of $m_\mathrm{min}$ below the minimum observed primary mass in the catalog, which is $\sim 6.4\msun$, to account for secondary BBH masses lower than that. We choose to fix the minimum black hole mass for both primary and secondaries to $5\msun$, 
similar to the inferred minimum mass in \citet{o3b_astro_dist}. The lack of observations of binaries with low primary mass make rate estimates in this region strongly model and prior dependent. We could be seeing signs of a decrease in merger rate from a ``lower mass gap'' between neutron star and BH masses, or we could be seeing fluctuations due to low-number statistics.  Either way we expect this to be resolved with future catalog updates. We also find no evidence for a sharp fall off in merger rate either following 
the pileup at ${\sim}35\,\msun$ -- expected if such a pileup was due to pulsational pair instability supernovae (PPISNe) -- or where the maximum mass truncation of the power law models 
are inferred. The lack of any high mass truncation, along with the peak at $\sim35\msun$ (significantly lower than expected from PPISNe) may pose challenges for conventional stellar 
evolution theory. This could be hinting at the presence of subpopulations that avoid pair instability supernovae during binary formation, but the confirmation of the existence of such subpopulations 
cannot be determined with the current catalog.

\begin{figure}[h]
    \includegraphics[width=\linewidth]{figures/q_distribution_plot.pdf}
    \caption{The marginal mass ratio distribution inferred with the B-Spline model (red), with 18 knots spaced linearly in $q$, from 0.05 to 1. The solid line shows the population predictive distribution (PPD), and the shaded region the 90\% credible interval. 
    We show the inferred PPD from the \textsc{PowerlawPeak} (blue) and \textsc{PowerlawSpline} (green) models from the LVK's GWTC-3 population analyses \citep{o3b_astro_dist}.}
    \label{fig:q_distribution}
    \script{q_distribution_plot.py}
\end{figure}

The marginal mass ratio distribution inferred by the B-spline model is shown in figure \ref{fig:q_distribution}.  These results suggest we may be seeing the first signs of departure from a simple power law behavior.  We find a potential plateau or decrease in the merger rate near equal mass ratios, as well as a broader tail towards unequal mass ratios 
than the power law based models find. Our results also suggest a shallower slope from $q\sim0.3$ to $q\sim0.7$, though uncertainty is larger in this region. The sharp decrease in rate just below 
$q\sim0.5$ is due to the minimum mass ratio truncation defined by $q_\mathrm{min}=\frac{m_\mathrm{min}}{m_1}$. When marginalizing over the primary mass 
distribution with a strong peak at $10\msun$, the mass ratio distribution truncates at $q\sim0.5$: the minimum mass, $5\msun$, 
divided by the most common primary mass, ${\sim}10\msun$.
